Deciding to Disclose: Pregnancy and Alcohol Misuse

### 2014-07-11 Background.tex
Pruning background

There has recently been an increased interest in the impact of binge
drinking, as a distinct pattern of consumption, with a wide variety
of negative outcomes reported by \citet{Maier2001}, although a significant
portion of their evidence base is drawn from animal studies which
augers for caution in generalising findings to humans. \citet{Strandberg-larsen2008}
explored links between binge drinking, and stillbirth, reporting a
statistically significant increase in risk associated with more than
three antenatal binge episodes. \citet{Sun2009} looked at seizure
disorders in children whose mothers binged during pregnancy. They
reported significantly greater risk of both neonatal seizures (\textasciitilde{}3
fold) and epilepsy (1.81 fold) associated with binge drinking between
11 and 16 weeks, but emphasised the exploratory nature of the results,
and need for replication. In terms of neurodevelopmental outcomes,
\citet{Streissguth1994} found a dose dependent association with scores
on timed word, and arithmetic tests in fourteen year olds with a stronger
association where bingeing occurred. A review by \citet{Henderson2007}
cautiously supports the contention that binge drinking has a neurodevelopmental
impact, but found no consistent support for adverse outcomes in pregnancy
(e.g. stillbirth, miscarriage, etc.) and note a paucity of studies
in the area. \citet{Meyer-Leu2011} considered the neonatal period,
finding that both moderate and binge drinking were associated with
an increased trend towards neonatal asphyxia. They also noted a large
number of contradictory findings and raising methodological concerns
about the studies reviewed by \citeauthor{Henderson2007}. \citet{Barr2006}
contend that binge drinking may also contribute to psychiatric issues
in the later life of offspring, although in this case their findings
are confined to individuals with \ac{FAS}, which may in itself be
a confounding factor, rather than indicating a directly causative
relationship between antenatal binge drinking and subsequent psychiatric
disorder in offspring.

@@

Given that alcohol consumption is thought to be underreported some
consideration must be given to the implications for midwifery practice,
in terms of eliciting more accurate self-reporting. \citet{Phillips2007}
present a qualitative account of factors influencing the disclosure
of substance misuse to midwives, identifying particularly the need
to build up a rapport, potentially over a number of appointments.
This was related to continuity of care, seen as necessary by both
midwives and women for building up a trust relationship, itself a
key component of facilitating disclosure. \citet{Stevens2002} looked
specifically at the process of transitioning to a caseloading model
of care provision in one midwifery practice, reporting that both practitioners
and women felt that this offered advantages in terms of long term
relationship building. Relationship building was also highlighted
by \citet{Kennedy2004} in a narrative investigation of midwifery
practice, where the subjects interpreted the midwife-woman dynamic
as about mutuality. \citeauthor{Kennedy2004} suggest that this arises
from a recognition that interactions in this context are about information
exchange, with the knowledge base of the woman as significant as that
of the midwife, rather than simply a unidirectionally didactic relationship.

\citet{Hunter2006} also focuses on much of midwifery as about relationship
building, suggesting that there is an insufficiently recognised emotional
labour component to practice. Observation and interview of a number
of midwives as they practiced suggested that many midwives effectively
took a mother type role to their patients, with implications around
the nature of information exchange that was able to take place. The
emotional labour component was also reported by \citeauthor{Stevens2002},
who suggest that this is more evident under a caseloading system, particularly
with challenging patients with complex needs. \citet{Todd1998} surveyed
midwives working in a hospital environment, as well as those working
in the community in a caseloading context, finding that community
practice appeared to provide more job satisfaction, but was challenging
to implement effectively because of limited resources. Community midwives
suggested that larger team sizes, and smaller caseloads would contribute
to a better realisation of the model. \citet{Farquhar2000} approached
the same question from the perspective of women, also finding that
faulty implementation hampered the expected benefits. They found that
those cared for under a team scheme, with much higher continuity of
care, reported that they had a better relationship with their midwives,
but were not more satisfied in general with their care. In contrast,
\citet{Biro2003} looked at a \ac{RCT} of team midwifery care versus
hospital care in Australia, finding a significantly higher level of
satisfaction under the team model, the distinction in this case may
lie in the different balance of team size to caseload size.

In terms of the impact of continuity of care on health, rather than
experiential outcomes, research is relatively sparse. \citet{Marks2003}
examined the impact of continuity of care on \ac{PND}, which has
similar features to alcohol in that it carries an associated stigma
that can act as a barrier to help seeking \citep{Dennis2006}. Based
on the results of a \ac{RCT}, they conclude that continuity of care
is not protective, in the sense of reducing rates or impacting onset,
but was very successful in supporting engagement with treatment. Echoing
this, a 2009 Cochrane Review by \citet{Hatem2009} found no significant
difference in incidence of \ac{PND}, but reported benefits in terms
of lower rates of \gls{episiotomies}, anaesthesia, and shorter hospital
stays, with higher satisfaction as found by \citeauthor{Biro2003}.
While research in this area does not specifically pertain to disclosure,
the general trend in results are suggestive when taken in concert
with studies emphasising the importance of relationship building as
key in fostering a disclosure friendly environment. Continuity of
care is generally regarded as improving patient experience, and leading
to better health outcomes in the wider medical arena \citep{VanWalraven2010},
but is clearly not a cost free endeavour, with particular concern
arising from the emotional cost \citep{Todd1998}, and increased rates
of `\gls{burnout}' in practitioners \citep{Sandall1997}.

>>>>

@@

Given that alcohol consumption is thought to be underreported some
consideration must be given to the implications for midwifery practice,
in terms of eliciting more accurate self-reporting. \citet{Phillips2007}
present a qualitative account of factors influencing the disclosure
of substance misuse to midwives, identifying particularly the need
to build up a rapport, potentially over a number of appointments.
This was related to continuity of care, seen as necessary by both
midwives and women for building up a trust relationship, itself a
key component of facilitating disclosure.
